---
title: "Population and Economic Effects of Storm Damage"
author: "RP"
date: "3/1/2021"
output: html_document
---
``` {r, echo = FALSE, cache = TRUE, message = FALSE}
if(!file.exists("StormData.csv.bz2")){
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "StormData.csv.bz2")
}

library(dplyr)
library(data.table)
library(ggplot2)
stormdata <- fread("StormData.csv.bz2")
```
# DATA PROCESSING

##Question 1
1. Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health? \Fatalities\ and \Injuries\
```{r}

pophealth <- stormdata %>% select(EVTYPE, FATALITIES, INJURIES)
## Combine the columns and filter out the zeroes
harmed <- pophealth %>% select(EVTYPE, FATALITIES, INJURIES) %>% mutate(HARMED = FATALITIES + INJURIES) %>% filter(HARMED != 0)
## Aggregate totals by the event type
harmed <- aggregate(cbind(FATALITIES, INJURIES, HARMED) ~ EVTYPE, data = harmed, FUN = sum)
## Arrange by total harmed, then by most fatalities
harmed <- arrange(harmed, desc(HARMED, FATALITIES))
## Plot the top ten?
harmed10 <- harmed[1:10,]
harmed10

```
```{r plotting}
ggplot(harmed10, aes(x = EVTYPE, y = INJURIES + FATALITIES)) + geom_bar(position = "stack", stat = "identity", fill = "2") + labs(title = "Effects of Storms on Population Health", x = "Storm Event", y = "Number of Victims")
```
#Question 2
2. Across the United States, which types of events have the greatest economic consequences? \PropertyDamage and \CropDamage
## Damage is in two columns: a "value" and an "exponent", where M = millions and K = thousands
```{r}

economy <- stormdata %>% select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
## Change letter suffixes in EXP to numeric multipliers of DMG
economy$CROPDMGEXP <- gsub("k", "1000", economy$CROPDMGEXP, ignore.case = TRUE)
economy$CROPDMGEXP <- gsub("m", "1000000", economy$CROPDMGEXP, ignore.case = TRUE)
economy$PROPDMGEXP <- gsub("k", "1000", economy$PROPDMGEXP, ignore.case = TRUE)
economy$PROPDMGEXP <- gsub("m", "1000000", economy$PROPDMGEXP, ignore.case = TRUE)
## Noticed some other modifiers while scrolling through the data...
## Find out what they are
table(economy$PROPDMGEXP)
economy$PROPDMGEXP <- gsub("h", "100", economy$PROPDMGEXP, ignore.case = TRUE)
economy$PROPDMGEXP <- gsub("b", "1000000000", economy$PROPDMGEXP, ignore.case = TRUE)
economy$PROPDMGEXP <- gsub("\\-|\\?|\\+|2|3|4|5|6|7|8|9", "1", economy$PROPDMGEXP, ignore.case = TRUE)
table(economy$CROPDMGEXP)
economy$CROPDMGEXP <- gsub("b", "1000000000", economy$CROPDMGEXP, ignore.case = TRUE)
economy$CROPDMGEXP <- gsub("\\?|2", "1", economy$CROPDMGEXP, ignore.case = TRUE)

economy$CROPDMGEXP <- as.numeric(economy$CROPDMGEXP)
economy$PROPDMGEXP <- as.numeric(economy$PROPDMGEXP)

##Combine the DMG and EXP columns

eco <- economy %>% mutate(PROP = PROPDMG * PROPDMGEXP, CROP = CROPDMG * CROPDMGEXP)
ecodmg <- aggregate(cbind(PROP, CROP) ~ EVTYPE, data = eco, FUN = sum)
ecodmg$TOTAL <- ecodmg$PROP + ecodmg$CROP
ecodmg <- arrange(ecodmg, desc(TOTAL))
ecodmg10 <- ecodmg[1:10,]
ecodmg10
```










