---
title: "Population and Economic Effects of Storm Damage"
author: "Ryan P"
date: "3/10/2021"
output:
  pdf_document: default
  html_document: default
  fig.caption: yes
---
## Synopsis:

  The data available from the U.S. National Oceanic and Atmospheric 
Administration's (NOAA) storm database was downloaded and analyzed to determine
what types of storms have the most effect on human health and cause the most
economic destruction. Through a little data processing and simple exploratory
bar graphs, some general conclusions can be made. Tornadoes are shown to be the 
leading cause of injury and death due to storms while floods cause the most 
economic damage to both property and agriculture.

  While a preliminary assessment of the data could be performed, it is extremely 
messy and requires extensive data processing far outside the scope of this 
assignment. Typos and variations in event types (e.g. "Thunderstorm WIND") make
it very hard to condense the data. The magnitude by which "Flood" and "Tornado"
surpass all others suggest that further processing will not change the 
conclusions made.

``` {r, echo = FALSE, cache = TRUE, message = FALSE}
if(!file.exists("StormData.csv.bz2")){
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.
              csv.bz2", "StormData.csv.bz2")
}

library(dplyr)
library(data.table)
library(ggplot2)
stormdata <- fread("StormData.csv.bz2")
```
```{r, echo = FALSE, message = FALSE}
library(dplyr)
library(data.table)
library(ggplot2)
```
## Data Processing for Question 1:

### Across the United States, which types of events  are most harmful with respect to population health?

```{r}
pophealth <- stormdata %>% select(EVTYPE, FATALITIES, INJURIES)
## Combine the columns and filter out the zeroes
harmed <- pophealth %>% select(EVTYPE, FATALITIES, INJURIES) %>% 
  mutate(HARMED = FATALITIES + INJURIES) %>% filter(HARMED != 0)
## Aggregate totals by the event type
harmed <- aggregate(cbind(FATALITIES, INJURIES, HARMED) ~ EVTYPE, data = harmed,
                    FUN = sum)
## Arrange by total harmed, then by most fatalities
harmed <- arrange(harmed, desc(HARMED, FATALITIES))
## Plot the top ten?
harmed10 <- harmed[1:10,]
harmed10

```

## Results for Question 1:

  A bar graph was produced showing the storm events that caused the top ten greatest amount of
injury and death to the US population. Tornadoes clearly lead with their destruction.
```{r figs, fig.cap = "Bar graph showing top ten destructive storm types for human health"}
ggplot(harmed10, aes(x = EVTYPE, y = INJURIES + FATALITIES)) + 
  geom_bar(position = "stack", stat = "identity", fill = "red") + 
  labs(title = "Effect of Storms on Population Health", 
       x = "Storm Type", y = "Number of Deaths and Injuries") +
  theme(axis.text.x=element_text(angle=33, hjust = 1))
```

## Data Processing for Question 2:

### Across the United States, which types of events have the greatest economic consequences?

```{r}
#Relevant columns are Property Damages and Crop Damages. 
#Both Damages have two columns: one with "value" and one an "exponent", 
#where M = millions and K = thousands
economy <- stormdata %>% select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
## Change letter suffixes in EXP to numeric multipliers of DMG
economy$CROPDMGEXP <- gsub("k", "1000", economy$CROPDMGEXP, ignore.case = TRUE)
economy$CROPDMGEXP <- gsub("m", "1000000", economy$CROPDMGEXP, ignore.case = TRUE)
economy$PROPDMGEXP <- gsub("k", "1000", economy$PROPDMGEXP, ignore.case = TRUE)
economy$PROPDMGEXP <- gsub("m", "1000000", economy$PROPDMGEXP, ignore.case = TRUE)
## Noticed some other modifiers while scrolling through the data...
## Find out what they are
table(economy$PROPDMGEXP)
##Make them disappear
##All weird ones are just changed to identity multiplier of "1"
economy$PROPDMGEXP <- gsub("h", "100", economy$PROPDMGEXP, ignore.case = TRUE)
economy$PROPDMGEXP <- gsub("b", "1000000000", economy$PROPDMGEXP, ignore.case = TRUE)
economy$PROPDMGEXP <- gsub("\\-|\\?|\\+|2|3|4|5|6|7|8|9", "1", economy$PROPDMGEXP
                           , ignore.case = TRUE)
table(economy$CROPDMGEXP)
economy$CROPDMGEXP <- gsub("b", "1000000000", economy$CROPDMGEXP, ignore.case = TRUE)
economy$CROPDMGEXP <- gsub("\\?|2", "1", economy$CROPDMGEXP, ignore.case = TRUE)

economy$CROPDMGEXP <- as.numeric(economy$CROPDMGEXP)
economy$PROPDMGEXP <- as.numeric(economy$PROPDMGEXP)
```
``` {r}
##Combine the DMG and EXP columns

eco <- economy %>% mutate(PROP = PROPDMG * PROPDMGEXP, CROP = CROPDMG * CROPDMGEXP)
ecodmg <- aggregate(cbind(PROP, CROP) ~ EVTYPE, data = eco, FUN = sum)
ecodmg$TOTAL <- ecodmg$PROP + ecodmg$CROP
ecodmg <- arrange(ecodmg, desc(TOTAL))
ecodmg10 <- ecodmg[1:10,]
ecodmg10
```

## Results for Question 2:

  A  bar graph was produced displaying the extent of damage caused by the top ten storm types. Floods clearly lead with their destruction.
```{r fig.cap = "Bar graph displaying the economic damage caused by the top ten storm types"}
dmg <- ggplot(ecodmg10, aes(EVTYPE, TOTAL))
dmg + geom_bar(stat = "identity", fill = "blue") + 
  theme(axis.text.x=element_text(angle=33, hjust = 1)) + 
  labs(title = "Total Economic Damage of Property and Crops", x= "Storm Type", 
       y = "Economic Damage ($)")
```







